---
title: "EF_Tutorial_1"
author: "Gillian"
output:
  pdf_document: default
  html_document: default
---
## Packages 
```{r, warning=F, message=F}
package_name <- "tidyverse"

# Check if the package is installed
if (!requireNamespace(package_name, quietly = TRUE)) {
  # If not installed, install the package
  install.packages(package_name)}

library(package_name, character.only = TRUE)
```

##Basic functions 
If you are not familiar with the function and want to get more info, add "?" at the front of the function to check its documentation (e.g. ?read_csv).

## Data importing 
```{r, warning=F, message=F}
library(tidyverse)
mydata <- read_csv("eaef21.csv") # load the dataset eaef21.csv into R 
#mydata <- read_csv("C:/Users/gilli/OneDrive/Documents/EF_tutorial_1/eaef21.csv")

#Sometimes your data is only available as a xlxs file.
#example <-read_exel(example)

#More info on importing various types of data https://github.com/rstudio/cheatsheets/blob/main/data-import.pdf

#Here is the chapter on data importing from "R for Data Sicence" https://r4ds.had.co.nz/data-import.html
```

#API 
#Can use API packages to access data without downloading it. This is very useful when dealing with large dataset or with ones that update regularly 
```{r}
#install.packages("wbstats")
library(wbstats)
gdp_data <- wb_data(country = c("AF","CA"), indicator = "NY.GDP.PCAP.CD")
```

```{r}
class(mydata)   # class will show us the type of object 'mydata'
mydata <- as.data.frame(mydata)   # change object to data.frame
class(mydata)
```

```{r, results = 'hide'}
head(mydata)     # shows the first 6 rows of your dataset
tail(mydata)    # shows the last 6 rows of your dataset
```
```{r}
summary(mydata) # gives you a simple summary of your dataset (not very nice looking)
```

```{r}
nrow(mydata)    # gives the number of rows
```

```{r}
ncol(mydata)    # gives the number of columns
```

The object mydata is a data.frame, i.e. a dataset with named columns. We can perform operations by rows and by columns with function apply.
```{r}
# this will apply function class to each column of the dataset
apply(mydata,     # the dataset to work on
      MARGIN = 2, # 1 to work by rows and 2 by columns
      FUN = mean) # the function to apply
```

```{r, eval=FALSE}
rm(gdp_data) # remove R object from environment
```

## Tidyverse
Tidyverse is a nice R package combining several packages for datamanagement and plotting, including \textbf{dplyr} \textbf{tidyr} and \textbf{ggplot2}. This package allows us to easily manipulate data.

```{r, warning=F, message=F}
if(!require(tidyverse)){install.packages("tidyverse")}
library(tidyverse)          # Package for data manipulation

# the following code gives same result as 
# mean(mydata$EARNINGS)
mydata %>% 
  summarize(mean(EARNINGS))
```

Create new varables with mutate.
```{r}
# create and add a new variable
mydata %>% 
  mutate(EDUCUNI = EDUCBA + EDUCMAST) %>%
  head(3)
```


Perform operations by clusters.
```{r}
# mutate/summarize by groups
mydata %>% 
  group_by(FEMALE) %>%
  summarize(mean(EXP))
```
```{r}
mydata %>% 
  group_by(MARRIED) %>%
  mutate(DHOURS = HOURS - mean(HOURS)) %>%
  head()
```

Perform more operations at the same time
```{r}
# you can also 'filter' rows and 'select' columns
mydata %>%
  mutate(AGE_D = ifelse(AGE <= 40, "below 40", "above 40")) %>%
  group_by(AGE_D) %>%
  summarise(mean(EARNINGS), mean(HOURS))
```


Pick only columns and observations we are interested into.
```{r}
# you can also 'filter' rows and 'select' columns
mydata %>% 
  filter(AGE > 40) %>%
  select(EARNINGS, EXP, FEMALE) %>%
  head()
```

In the above example we are using a pipe "%>%". This comes from the package magrittr and is compatible with all tidyverse packages. 

Here we complete the same but ussing the classic R approach
```{R}
# classic R
classic <- filter(mydata, AGE > 40)
classic1 <- select(classic,c(EARNINGS, EXP, FEMALE))
classic2 <- head(classic1)
print(classic2)
```

Now, let’s put that in practice!
Can you write some code to compare the average male and female earnings for above 40 year olds with a masters 
```{r}
avg_earnings <- mydata %>%
  filter(AGE > 40, EDUCMAST == 1) %>%
  group_by(MALE) %>%
  summarize(AverageEarnings = mean(EARNINGS, na.rm = TRUE))

head(avg_earnings)
# 0 represents women
# 1 represents men
```

# Univariate analysis

We can also easily plot our data. We can use histograms for numerical variables.
```{r}
mydata %>% 
  ggplot(aes(EARNINGS)) +
    geom_histogram(bins = 35, 
                   color = "black", 
                   fill = "white") +
    ylab("") + 
    xlab("") +
    ggtitle("Histograms of Earnings (Frequency)")  
```

We can add a density line, but we need to show histograms as probabilities.
```{r}
mydata %>% 
  ggplot(aes(EARNINGS)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 100, 
                   color = "grey30", 
                   fill = "white") +
        geom_density(alpha = .2, 
                     fill = "antiquewhite3") +
    ylab("") + 
    xlab("") +
    ggtitle("Histograms of Earnings (Density)")  
```

For qualitative analysis you can try for yourself geom_bar() and geom_count().

## Univariate regression
```{r}
mod0 <- mydata %>%
  lm(log(EARNINGS) ~ EXP, data = .)
  
mod0 %>% summary()
```


# Multivariate analysis
We can now start with multiple linear regression model.
```{r}
mod1 <- mydata %>%
  lm(log(EARNINGS) ~ AGE + EXP + 
       EDUCPROF + EDUCPHD + 
       EDUCMAST + EDUCBA + 
       EDUCAA + EDUCHSD + 
       EDUCDO, data = .)
  
mod1 %>% summary()
```


